\subsection{Oracle-Guided Inductive Synthesis}
\label{sec:ogis}

\textit{\Glsfmtlong{ogis}~(\Gls{ogis})} is an approach to program synthesis
where the synthesizer is split into two components: the \textit{learner} and the
\textit{oracle}.
The two components communicate in an iterative \textit{query/response} cycle, as
shown in Figure~\ref{fig:ogis}.
The learner implements the search strategy to find the program and is
parameterized by some form of semantic and/or syntactic specifications
(see~\ref{sec:specifications}).
The usefulness of the oracle is defined by the type of queries it can handle and
the properties of its responses. The characteristics of these components are
typically imposed by the application.

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    [semithick, >=stealth, auto,
     rectangular/.style={rectangle, draw, rounded corners, text width=4cm,
       align=center, minimum size=1.5cm},
     spherical/.style={circle, draw, text width=2cm, align=center}]

    \node [rectangular] (S)  {Learner};
    \node [left=1.95cm of S, align=center] (I) {Specification\\and/or Syntactic Bias}
      edge [->] (S);
    \node [below=of S, align=center] {Program $p$\\or Fail}
      edge [<-] (S);
    \node [spherical] (V)  [right=3cm of S] {Oracle}
      ([yshift=0.2cm]S.east) edge [->, bend left]  node        {Query}    ([yshift=0.2cm]V.west)
      ([yshift=-.2cm]S.east) edge [<-, bend right] node [swap] {Response} ([yshift=-.2cm]V.west);
  \end{tikzpicture}
  \caption{\Acrlong{ogis}. Adapted from
    \protect\citeauthor{Jha:2017:TFS}~\protect\cite{Jha:2017:TFS}.}
  \label{fig:ogis}
\end{figure}

Typical queries and response types are some of the following~\cite{Jha:2017:TFS}:

\begin{itemize}
\item \textit{Membership queries}, where given an input-output example $x$ the oracle
  responds with the answer to whether $x$ is positive or not.
\item \textit{Positive (resp. negative) witness queries}, where the oracle
  responds with a positive (resp. negative) input-output example, if it can find
  any, or $\bot$ otherwise.
\item \textit{Counterexample queries}, where given a candidate program $p$ the
  oracle responds with a positive input-output counterexample that $p$ does not
  satisfy, if it can find any, or $\bot$ otherwise.
\item \textit{Correctness queries}, where given a candidate program $p$ the
  oracle responds with the answer to whether $p$ is correct or not. If it is not,
  the oracle responds with a positive input-output counterexample.
\item \textit{Verification queries}, where given program $p$ and specification
  $\phi$ the oracle responds with the answer to whether $p$ satisfies $\phi$ or
  not, or $\bot$ if it cannot find the answer.
\item \textit{Distinguishing input queries}, where given program $p$ and a set
  $X$ of input-output examples that $p$ satisfies, the oracle responds with a
  new counterexample $x$ to $p$ such that another program $p'$ exists that
  satisfies both $x$ and all the other examples in $X$.
\end{itemize}

An \gls{ogis} system responding to counterexample queries corresponds to the
\textit{\gls{cegis}} system, introduced by
\citeauthor{Solar-Lezama:2008}~\cite{Solar-Lezama:2008} in the context of the
SKETCH synthesizer. Correctness oracles are more powerful than counterexample
oracles because they are guaranteed to return a counterexample if the program is
not correct, where the counterexample oracles might not.

The concept of \gls{ogis} was introduced by
\citeauthor{Jha:oracle:2010}~\cite{Jha:oracle:2010} as a generalization of
\gls{cegis} when they applied this idea to a \gls{pbe} synthesizer based on
distinguishing inputs in order to deobfuscate malware and to generate
bit-manipulating programs.
Jha et al. further developed this idea by presenting a new theoretical framework
for inductive synthesis~\cite{Jha:2017:TFS}.

In general, the higher the capabilities of the oracles, the more expensive they
are to run. Distinguishing oracles are (typically) not as strong as
counterexample or correctness oracles because the returned counterexample is not
necessarily positive. To understand why they might be effective tools we can
turn to the Bounded Observation Hypothesis~\cite{Solar-Lezama:2008}, which
asserts that ``an implementation that works correctly for the common case and
for all the different corner cases is likely to work correctly for all inputs.''

In a setting where the synthesizer is allowed to interact with the user, we
could see the users take the role of the oracles.
However, the interesting cases are the ones where the ratio between the amount
of work the users are given and the information given to the synthesizer is
minimized.
A system that frequently queries the users for correctness checks would probably
feel very cumbersome.
On the other hand, a system that queries for membership or positiveness checks
might be more realistic, as usually the user has an idea of what sort of
examples fit their desired model.

