\section{Whole Encoding}
\label{sec:whole-encoding}

In the previous approach the workload was split between the enumerator and the
solver: the enumerator would exhaustively search for all combinations (with
replacement) of the library of components in order of increasing size and
pass them one by one to the solver.
In turn, the solver would then verify if a program existed that satisfied the
input-output examples using each of the given components exactly once.
However, the number of combinations of a given size grows exponentially.
One wonders if it is possible to free the enumerator by pushing as much workload
as possible to the solver.

In the next approach we reduce the enumerator to query the solver with a single
number $n$ (plus the input-output examples). Instead of drawing components from the
library and passing them to the solver, the enumerator now queries if there is 
any program of exactly $n$ components that satisfies the examples. Thus, the
solver is now parameterized by the library of components $F$. This process is
repeated for increasing values of $n$ until a program is found.

\subsection{Program Formula}
\label{sec:program-formula-whole}

In this section, we adapt the location-based encoding from the previous section
in order to find a program with the exact number of allowed components, $n$.
Just as before, we will have a set $I$ of input variables, an output variable
$o$, a set $C$ of constant variables, a set $L$ of location variables, a set $R$
of return variables, and a set $P$ of formal parameter variables, with some
changes, as we will see next.

In this scenario we do not know how many times each component will be used in
the synthesized program.
This means we must have some way to model the choices of which components get
picked and which do not.
For this reason we introduce a new set of integer-valued variables $A = \{a_1,
a_2, \ldots, a_n\}$, which we will refer to as the \textit{activation}
variables.
The activation variables have values in the set $\{1, 2, \ldots, |F|\}$, with
the interpretation that if we have $a_i = k$, then component $f_k$ will be the
$i$-th component in the program (the one appearing in line $|I| + |C| + i$).
For the same reason, this time we do not assign return variables to each
component.
Instead, since we know the number $n$ of components we are aiming for, we have
one return variable $r_i$ associated with each $a_i$, for $i = 1, \ldots, n$.
These components will have their locations fixed beforehand: variable $r_i \in
R$ will be assigned to location $l_{r_i} = |I| + |C| + i$.

\begin{example}
  Consider the program in Figure~\ref{fig:first-name-ssa}.
  Assuming that component \lstinline{Concat} corresponds to $k = 1$,
  \lstinline{Index} to $k = 2$, and \lstinline{Substr} to $k = 3$, then we would
  have $a_1 = 2 $, $a_2 = 3$, and $a_3 = 1$.
  We would also have $l_{r_1} = 5$, $l_{r_2} = 6$, and $l_{r_3} = 7$.
\end{example}

Since the values of the activation variables $x \in A$ will only be known after
solving the formula, we do not know apriori the concrete types of the return
variables $x \in R$.
We circumvent the problem by augmenting the type of each return variable to be
the union of all possible return types.
In our case, this implies that all return variables are either of type
\lstinline{Text}, or of type \lstinline{Integer}.

The same happens for the types of the formal parameter variables $x \in P$.
Moreover, we do not know exactly how many formal parameter variables to create
because we do not know the arities of the components that will be picked for
each line.
We know, however, the maximum arity $m$ of all components.
Hence, there is an upper bound on the number of formal parameter variables we
might need, namely $n * m$.
Thus, we introduce variables $p_{ij} \in P$, for $i = 1, 2, \ldots, n$, and
$j = 1, 2, \ldots, m$, where each $p_{ij}$ is the $j$-th parameter of $f_{a_i}$,
the component activated by $a_i$.
We also augment their type to be the union of all possible formal parameter
types, plus a \lstinline{Null} type, inhabited by a single value,
\lstinline{null}, to indicate the absence of value.
A variable $p_{ij} \in P$ will be \lstinline{null} if and only if the arity of
component $f_{a_i}$ is less than $j$, and consequently will be ignored by
$f_{a_i}$ if that is the case.
Finally, we perform the same \lstinline{Null} type augmentation to the location
variables $x \in L$, because it makes no sense in having a location number if a
variable is null.

\begin{figure}
  \begin{lstlisting}[frame=tlrb,mathescape=true]
    prog($i_0$, $i_1$):
      $c_0$ = ?
      $c_1$ = ?
      $r_1$ = $f_{a_1}$($p_{11}$, $p_{12}$, $p_{13}$)
      $r_2$ = $f_{a_2}$($p_{21}$, $p_{22}$, $p_{23}$)
      $r_3$ = $f_{a_2}$($p_{31}$, $p_{32}$, $p_{33}$)
  \end{lstlisting}
\caption{Symbolic representation of a program with two inputs, $i_0, i_1 \in I$,
  two constant variables $c_0, c_1 \in C$, three return value variables $r_i \in
  R$, for $i = 1, \ldots, n$, and nine formal parameter variables $p_{ij} \in
  P$, for $i = 1, \ldots, n$, $j = 1, \ldots, m$, with $n = m = 3$.
  The question marks '?' are values to be found by the solver.}
\label{fig:whole-encoding-prog}
\end{figure}

\begin{example}
  Consider program~\ref{fig:first-name-ssa} and its corresponding symbolic
  skeleton from Figure~\ref{fig:whole-encoding-prog}.
  We have three components: \lstinline{Concat}, \lstinline{Index}, and
  \lstinline{Substr}.
  Components \lstinline{Index} and \lstinline{Substr} have arity 3,
  which is the largest among the three components.
  Thus, we have $n = m = 3$. This means we have nine variables $p_{ij} \in P$.
  Variables $p_{1j}$, with $j = 1, \ldots, 3$ are the formal parameters of
  $f_{a_1}$, which is component \lstinline{Concat}.
  However, \lstinline{Concat} only takes two parameters.
  These will be $p_{11}$ and $p_{12}$, meaning that $p_{13}$ will take value
  \lstinline{null}.
\end{example}

\subsubsection{Well-Formedness Constraint}
\label{sec:whole-well-formedness-constraint}

As in the setwise encoding (Section~\ref{sec:setwise-encoding}), each input
variable $x \in I$ is assigned a distinct location $l_x \in L$ from 1 to $|I|$,
and each constant variable $x \in C$ is assigned a distinct location from $|I| +
1$ to $|I| + |C|$.
As said in the previous section, this time we do the same for each return
variable $x \in R$, assigning to each a location from $|I| + |C| + 1$ to
$|I| + |C| + |R|$.
The upper bound on the location (if non-null) of each parameter variable $p_{ij}
\in P$ is now statically known to be $|I| + |C| + i$, so now our well-formedness
constraint $\psi{}_{\mathtt{wfp}}$ is simply:
%
\[
  \psi{}_{\mathtt{wfp}}(L,C,I,P) =
  \bigwedge_{i = 1, \ldots, n}\bigwedge_{j = 1, \ldots, m}
  (l_{p_{ij}} \neq \mathtt{null} \implies 1 \leq l_{p_{ij}} < |I| + |C| + i)
\]
%
\noindent
Compared to the well-formedness constraint of the setwise encoding, this
constraint is arguably simpler due to the fact that the lines of the return
variables are fixed apriori, and because we are not doing any type checking.

\begin{example}
  Consider the symbolic skeleton from Figure~\ref{fig:whole-encoding-prog}.
  We have:
  % 
  \begin{align*}
    &l_{i_0} = 1 \quad l_{i_1} = 2 \\
    &l_{c_0} = 3 \quad l_{c_1} = 4 \\
    &l_{r_1} = 5 \quad l_{r_2} = 6 \quad l_{r_3} = 7
  \end{align*}
  % 
  Also, the well-formedness constraint is the conjunction of the following:
  % 
  \begin{align*}
           l_{p_{11}} \neq \mathtt{null} \implies 1 \leq l_{p_{11}} < 5
    \qquad l_{p_{12}} \neq \mathtt{null} \implies 1 \leq l_{p_{12}} < 5
    \qquad l_{p_{13}} \neq \mathtt{null} \implies 1 \leq l_{p_{13}} < 5 \\
    %
           l_{p_{21}} \neq \mathtt{null} \implies 1 \leq l_{p_{21}} < 6
    \qquad l_{p_{23}} \neq \mathtt{null} \implies 1 \leq l_{p_{23}} < 6
    \qquad l_{p_{22}} \neq \mathtt{null} \implies 1 \leq l_{p_{22}} < 6 \\
    %
           l_{p_{31}} \neq \mathtt{null} \implies 1 \leq l_{p_{31}} < 7
    \qquad l_{p_{32}} \neq \mathtt{null} \implies 1 \leq l_{p_{32}} < 7
    \qquad l_{p_{33}} \neq \mathtt{null} \implies 1 \leq l_{p_{33}} < 7
  \end{align*}
\end{example}
%
\subsubsection{Functional Constraint}

Similarly to the last encoding, we have the following concerns regarding the
semantics of the program:
\begin{enumerate*}[(1)]
\item \label{itm:whole-fc-one} relating return variables to the values of their
components and their parameters;
\item \label{itm:whole-fc-two} value sharing between variables with the same
location; and
\item \label{itm:whole-fc-three} effectively mapping the inputs $x \in I$ to the
output $o$.
\end{enumerate*}

Constraint $\psi{}_{\mathtt{spec}}$, which guarantees
constraint~\ref{itm:whole-fc-one}, should be similar to the one from
section~\ref{sec:functional-constraint}, but we will have to pay attention to
some special points.
First, we have to make sure that, for every component, we relate the correct
number of formal parameters to the return value:\footnote{Recall that we may
have more formal parameter variables than we may need because we do not know
apriori which component goes on which line.}
%
\[
  \psi{}_{pnum}(A, P, R) =
  \bigwedge_{i=1,\ldots,n}\bigwedge_{k=1,\ldots,|F|}
  (a_i = k \implies \phi{}_{f_{k}}(p_{i1}, p_{i2}, \ldots, p_{ij_k}, r_i))
\]

\noindent
In the constraint shown above, $j_k$ is used as shorthand for the arity of
component $f_k$.
Second, we should ensure that formal parameter variables that are in excess
have value \lstinline{null}:
%
\[
  \psi{}_{null}(A, P) =
  \bigwedge_{i=1,\ldots,n} \bigwedge_{k=1,\ldots,|F|}
  (a_i = k \implies \bigwedge_{j=j_k+1,\ldots,m} p_{ij} = \mathtt{null})
\]

\noindent
Finally, we need to constrain the formal parameters to have the correct type,
because the specifications $\phi{}$ of the components are typed.
An example might, perhaps, be the best way to show how this constraint, which we
will call $\psi{}_{type}$, can be materialized.

\begin{example}
  Consider the symbolic skeleton from Figure~\ref{fig:whole-encoding-prog}, and
  suppose that we are working with components
  $f_1 = \mathtt{Concat}$, $f_2 = \mathtt{Index}$, and $f_3 = \mathtt{Substr}$.
  We will use the predicates $isstring$ and $isint$ to mean that a variable must
  be of type \lstinline{Text}, or \lstinline{Integer}, respectively.
  In this concrete case, the following constraint is needed to ensure that the
  type signatures of these components are satisfied:
  % 
  \begin{align*}
    \psi{}_{type}(A, P)
    &= \bigwedge_{i=1,\ldots,n}
      (a_i = 1 \implies (isstring(p_{i1}) \wedge isstring(p_{i2}))) \\
    &\wedge \bigwedge_{i=1,\ldots,n}
      (a_i = 2 \implies (isstring(p_{i1}) \wedge isstring(p_{i2}) \wedge isint(p_{i3}))) \\
    &\wedge \bigwedge_{i=1,\ldots,n}
      (a_i = 3 \implies (isstring(p_{i1}) \wedge isint(p_{i2}) \wedge isint(p_{i3})))
  \end{align*}
\end{example}

\noindent
Constraint $\psi{}_{\mathtt{spec}}$ is then just the conjunction of the three
previous constraints:
%
\[
  \psi{}_{\mathtt{spec}}(A, P, R) =
  \psi{}_{pnum}(A, P, R) \wedge \psi{}_{null}(A, P) \wedge \psi{}_{type}(A, P)
\]

The dataflow properties of constraint~\ref{itm:whole-fc-two} are encoded in a
similar fashion as in Section~\ref{sec:functional-constraint}, with the added
constraint that every formal parameter variable with \lstinline{null} value must
also have a \lstinline{null} location:
%
\begin{align*}
  \psi{}_{flow}(C,I,P,R)
  &=\bigwedge_{i=1,\ldots,n}
    \bigwedge_{j=1,\ldots,m}
    \bigwedge_{\substack{x \in I \cup C \cup R}}
    (l_{p_{ij}} = l_x \implies p_{ij} = x) \\
  &\wedge\bigwedge_{i=1,\ldots,n}
    \bigwedge_{j=1,\ldots,m}
    (l_p = \mathtt{null} \implies p = \mathtt{null})
\end{align*}
%
\noindent
Again, we need to ensure that every return variable is used (as an actual
parameter or as the output of the program).
This property is encoded as follows:
%
\begin{align*}
  \psi{}_{out}(o, P, R)
  =\bigwedge_{k=1,\ldots,n-1}
    \bigvee_{i=k+1,\ldots,n}
    \bigvee_{j=1,\ldots,m}
    (l_{r_k} = l_{p_{ij}})
  \wedge (r_n = o)
\end{align*}
%
\noindent
Formula $\psi{}_{out}$, along with $\psi{}_{flow}$, ensures that
constraint~\ref{itm:whole-fc-three} is satisfied.
We also need to ensure that every input variable appears as an actual parameter
of some component, a property encoded by the following formula:
%
\begin{align*}
  \psi{}_{in}(I, P)
  =\bigwedge_{x \in I}
    \bigvee_{i=1,\ldots,n}
    \bigvee_{j=1,\ldots,m}
    (l_{x} = l_{p_{ij}})
\end{align*}
%
The functional constraint is obtained by conjoining the previous constraints
with the well-formedness constraint $\psi{}_{\mathtt{wfp}}$,
again wrapping the formal parameter and return value variables $x \in P \cup R$
under an existential quantifier:
%
\begin{align*}
  \psi{}_{prog}(A,L,C,I,o) &= \exists P,R\ldotp
  (\psi{}_{\mathtt{wfp}}(L, P) \wedge
  \psi{}_{spec}(A, P, R) \wedge
  \psi{}_{flow}(I, C, P, R)\\&\wedge 
  \psi{}_{out}(o, P, R) \wedge
  \psi{}_{in}(I, P))
\end{align*}

\subsubsection{Full Constraint}
\label{sec:whole-full-constraint}

The functional constraint $\psi{}_{prog}$ encodes the space of well-formed
programs that satisfy a specific input-output example $(I, o)$.
To obtain a formula that ranges over all input-output examples, we just make a
conjunction over the set $E$:
%
\[
  \Psi{}=\exists{A,L,C}\ldotp\bigwedge_{(I, o) \in E}\psi{}_{prog}(A,L,C,I,o)
\]
%
\noindent
A model of this formula is an assignment to the variables $x \in A \cup L \cup C$,
which can then be used to reconstruct a program that satisfies the input-output
examples $E$ using only $n$ components.