\chapter{Conclusion and Future Work}
\label{chap:conclusion}

In this work, we tackle the problem of synthesizing OutSystems expressions from
examples, focusing on expressions that manipulate integers and text.
OutSystems sells solutions to help building enterprise-grade applications
swiftly and with a great degree of automation.
Thus, we were interested in an ``push-button''-style approach that would be
performant, and could generalize from a small number of examples.
We surveyed the state of the art in program synthesis, and implemented two
component-based \gls{pbe} synthesizers -- setwise and whole.
Both synthesizers employ a mixture of constraint solving with basic enumerative
search, differing on the amount of work they put on the constraint solving
phase.
We benchmarked both synthesizers and compared them to SyPet~\cite{Feng:2017:CSC}
with two different configurations: one where all constants needed to solve the
instances are given as new components, and another that simulates
user-provided constants by only allowing the constants needed for each
particular instance.

The setwise synthesizer can consistently synthesize programs up to four lines
that satisfy the given input-output examples, but only manages to match the user
intent on programs with size up to three lines.
Still, it manages to be competitive with SyPet, even when we configured the
latter for a scenario simulating user-provided constants.
However, the enumerator of the setwise synthesizer is very simple, and it would
be interesting to see how more elaborate approaches, such as stochastic search,
would fare.
On the other hand, the whole synthesizer fails to produce good results on both
fronts in all but the most trivial instances (programs with one line).
However, additional variations and refinements of the encodings should be
tested.
It would also be interesting to test configurations with user-provided inputs.
Another interesting scenario, probably involving extensions to the synthesizers
includes trying to figure out from the input-output examples which constants the
program might use.
Moreover, it would be interesting to see how our synthesizers compare to a
synthesizer implemented using the PROSE
framework.\footnote{\url{https://microsoft.github.io/prose/}}

Both approaches use encodings which size scales linearly with the number of
input-output examples.
While not ideal, it is not unacceptable given that we are interested in
offloading as much work as possible from the user, and so the approach should
generalize with a small number of examples.
Given the experimental results, tending towards approaches more reliant on
general-purpose constraint solving might not be the best way to tackle this kind
of problems.
On the other hand, it would be interesting to explore how solvers specialized
for synthesis, or approaches with a tighter integration with the underlying
solver might fare.
In fact, some approaches in the literature apply this
idea~\cite{Feng:2018:PSU,Abate:2018:CMT}.
Moreover, it should be explored whether this approach scales when a larger
library of components is used, especially components that do not have a direct
encoding in \gls{smt} (and, in turn, could be more difficult, or even impossible
to synthesize).
It would be particularly interesting to support if-then-else expressions, along
with frequently used predicates over Integers and Texts.
